### 2023-03-08 paper-thesis/chapters/11-variable-capture.tex
Work on interpreter text + other work on text

Almost all compilers that typecheck dependently typed languages use de Bruijn representation for variables. Using de Bruijn indices in statix is possible, but sacrifices a lot. Many editor services (renaming, go to definition) require \verb|.ref| annotations to be set on names, and this is not possible if the names are no longer a part of the AST.

@@

Another solution that was attempted was having a pre-analysis transformation that gives each variable a unique name. This doesn't work for a variety of reasons. First of all, it doesn't actually solve the problem. Names can be duplicated during beta reduction of terms, so we still don't have the guarantee that each variable has a unique name. Furthermore, this changes the names in the AST, so it breaks editor services such as renaming.

\subsection{Renaming terms dynamically}

Anytime that we introduce a new name in a type, we could check if the name already exists in the environment, and if it does, choose a different unique name. This approach is possible but tedious to implement in Statix. It requires a new relation to traverse through the type and rename. This is also an inefficient solution, as many traversals of the type are needed. This was successfully implemented, but we chose against using it in favor of the solution presented in the next section.

@@

Because the expression above, with \verb|ScopedName|s, is not particularly readable, we add a new post-analysis Stratego pass that converts the \verb|ScopedName|s to ticked names. For example, the above would be transformed to:

>>>>

Almost all compilers that typecheck dependently typed languages use de Bruijn representation for variables~\cite{lean}. Using de Bruijn indices in statix is possible, but sacrifices a lot. Many editor services (renaming, go to definition) require \verb|.ref| annotations (which specify which other name a certain name refers to) to be set on names, and this is not possible if the names are no longer a part of the AST and the scope graph.

@@

Another solution that was attempted was having a pre-analysis transformation that gives each variable a unique name. This doesn't work for a variety of reasons. First of all, it doesn't actually solve the problem. Names can be duplicated during beta reduction of terms, so we still don't have the guarantee that each variable has a unique name. Furthermore, this changes the names in the AST, so it breaks editor services in the same was as De Bruijn Indices did.

\subsection{Capture-avoiding substitution}

Anytime that we introduce a new name in a type, we could check if the name already exists in the environment, and if it does, choose a different unique name. This approach is called capture-avoiding substitution by renaming~\cite{capture_avoiding_sub}. This is possible but tedious to implement in Statix. It requires a new relation to traverse through the type and rename. This is also an inefficient solution, as many traversals of the type are needed. This was successfully implemented, but we chose against using it since we found a better solution.

@@

Because the expression above, with \verb|ScopedName|s, is not particularly readable, we add a new post-analysis Stratego pass (an AST transformation that runs directly after type-checking) that converts the \verb|ScopedName|s to ticked names. For example, the above would be transformed to:

### 2023-03-08 paper-thesis/chapters/20-adding-booleans.tex
Work on interpreter text + other work on text

>>>>

It is important that our implementation is easily extensible since real-world dependently typed languages have complex features such as inductive data types and term inference, that make the language a lot easier to use. We will first show how to add support for these relatively simple features, and then use the same techniques to add support for more complex features in the following chapters.

### 2023-03-08 paper-thesis/chapters/24-backend.tex
Work on interpreter text + other work on text

However, this big advantage is also a disadvantage: Dynamic semantics would usually be defined in Stratego. We need to find a way to re-use the definition from Statix, since we don't want to repeat the definition. Luckily, Statego has a way to evaluate Statix relations on terms.

\todo{finish this after implementing}
- discuss efficiency (reference "Optimizing first class pattern matching")

>>>>

However, this big advantage is also a disadvantage: Dynamic semantics would usually be defined in Stratego. We need to find a way to re-use the definition from Statix, since we don't want to repeat the definition of the dynamic semantics.

\subsection{Implementation}

Luckily, Stratego has a strategy built in that runs a Statix predicate on a term: \verb|stx-evaluate|. Using this strategy, the actual implementation of our interpreter in Stratego is just one line: Match an expression \verb|e| and run the functional predicate \verb|interpret| from the \verb|main.stx| file on it.

\begin{lstlisting}
interpret : e -> <stx-evaluate(|"main", "main!interpret")> [e]
\end{lstlisting}

Then, in the \verb|main.stx| file we define \verb|interpret| to be beta reducing the expression in an empty scope.

\begin{lstlisting}
interpret(e) = betaReduce((sEmpty(), e)).
\end{lstlisting}

These two lines are the only required changes to implement an interpreter\footnote{Other than changing the project config to define a button in the Spoofax UI that when pressed runs the interpret strategy on a given term}. This shows that it is very straight-forward to implement an interpreter.

\subsection{Discussion of defining in Statix versus Stratego}

An advantage of implementing the dynamic semantics in Statix instead of Stratego is that we can use scope graphs as a part of our dynamic semantics rules.

A disadvantage is that Stratego is specifically optimized for these transformations~\cite{fcpmc}. Therefore, the resulting interpreter will be slower when implemented in Statix. \todo{comparison}
